---
title: "Power analysis in Developmental Psychology"
author: "Gonzalo García-Castro"
date: "25/07/2019"
output: ioslides_presentation
css: "2019-07-25_workstats_power.css"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	include = TRUE
)
```

```{r load_packages, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(magrittr)
library(dplyr)
library(tidyr)
library(forcats)
library(broom)
library(purrr)
library(ggfortify)
library(ggplot2)
library(ggthemes)
library(knitr)

alpha <- 0.05
```

# Some context...

## Statistical inference

The goal of every statistical inference is to:

* Make the decission most **supported by data**
* Accurately represent our **uncertainty** on such decission

```{r decissions, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
data_decission <- data.frame(
  experiment = c(rep("More confidence", 10000),
                 rep("Less confidence", 60)),
  group = c(rep("Monolinguals", 5000),
            rep("Bilinguals", 5000),
            rep("Monolinguals", 30),
            rep("Bilinguals", 30)),
  score = c(rnorm(n = 5000, mean = 0, sd = 1),
            rnorm(n = 5000, mean = 5, sd = 1),
            rnorm(n = 30, mean = 0, sd = 2),
            rnorm(n = 30, mean = 5, sd = 2))
)

summary_decission <- data_decission %>%
  group_by(experiment, group) %>%
  summarise(n = n(), mean = mean(score))

ggplot(data_decission, aes(x = score, fill = group)) +
  geom_density(alpha = 0.5, bw = 1.5) +
  geom_vline(data = summary_decission, aes(xintercept = mean)) +
  geom_label(data = summary_decission, aes(x = mean, y = 0.1),
             label = paste("n =", summary_decission$n), fill = "white") +
  labs(x = "Score", y = "Probability density", fill = "Group",
       title = "Same decission, different confidence",
       subtitle = "Red lines: Mean") +
  scale_fill_few(palette = "Medium") +
  theme(
    text = element_text(size = 12),
    axis.text = element_text(colour = "black"),
    legend.position = "none"
  ) +
  facet_wrap(~experiment, ncol = 1)


```

## NHST | One path toward statistical inference

**Null Hypothesis Significance Testing (NHST)**

Specify a null hypothesis ($H_0$) and an alternative hypothesis ($H_1$). 

Test the $H_0$ and decide whether to reject it or not:

1. Asume $H_0$ is true.
2. Calculate a statistical value:
    a) That reflects the parameter we are interested in (e.g. *t*-value for differences in means or *F*-value for variance explained by factor)
    b) For which we know its probabilistic distribution at the population level (e.g. Student's *t* or Snedecor's *F*)
3. If the probability associated to our statistic (according to the distribution) is lower than our significance criterion ($\alpha$), we reject the $H_0$: our data is very unlikely to have been generated under such scenario.

## Null Hypothesis Significance Testing

Specify a null hypothesis ($H_0$) and an alternative hypothesis ($H_1$). 

Test the $H_0$ and decide whether to reject it or not:

1. Asume the **null hypothesis** is true.
2. Calculate a statistical value:
    a) That reflects the parameter we are interested in (e.g. *t*-value for differences in means or *F*-value for variance explained by factor)
    b) For which we know its probabilistic distribution at the population level (e.g. Student's *t* or Snedecor's *F*)
3. If the probability (*p*) associated to our statistic (according to the distribution) is lower than our **significance** criterion ($\alpha$), we reject the null hypothesis: our data is very unlikely to have been generated under such scenario.

## *p*-values

*p*-values in the context of NHST indicate the probability of our data (i.e. statistic of contrast) under the null hypothesis.

We compare the *p*-value we have against our significance criterion.

Significance criterion ($\alpha$): how unlikely must our data be to be considered too unlikely to have been generated under the null hypothesis.

<center>If *p*-value < $\alpha$ $\to$ Reject $H_0$</center>

## *p*-values

Let's take a look at Daniel Laken's shiny app for exploring how *p*-values behave under some conditions: [[link]](https://lakens.shinyapps.io/p-curves/).

* Under $H_0$, all *p*-values are equally likely.
* Under $H_1$, lower *p*-values are more likely.

## Two types of incorrect decisions

Statistical inference is **probabilistic**. We cannot know if our decission to reject (or not to reject) the null hypothesis is correct. If we knew it, statistical inference would be unnecessary.

We can make two types or errors:

* **False positives** (Type I error, $\alpha$): There is not an effect, we found an effect
* **False negatives** (Type II error, $1- \beta$): There is an effect, we did not find an effect.

Every time we reject $H_0$, it could be a Type I error.
Every time we do not reject $H_1$, it could be a Type II error.

## Controlling our error rates | Type I error rate

Specify the cost of each error, and adjust the rest of parameters to them.

Assume there is **no effect**. The probability of (incorrectly) rejecting $H_0$ is $\alpha$.

If we set $\alpha$ very low, only very unlikely *p*-values will lead to the rejection of $H_0$
If we set $\alpha$ very high, we could reject $H_0$ when *p*-values are quite high.

Usually false positives are considered more **costly**, so we usually fix $\alpha$ at a very low level (e.g. 0.05), and adjust the rest of parameters.

## Controlling our error rates | Type I error rate

I need to get an NBA player. How can I select one person to make my best guess?

[Article](https://www.si.com/vault/2015/07/05/106084744/larger-than-real-life) by Pablo S. Torre for *Sports Illustrated*: there is a .17 probability that, if you are a 7 feet high American man, you are an NBA player.


## Controlling our error rates | Type II error rate

Assume there **is** an effect. The probability of (incorrectly) not rejecting $H_0$ is $\beta$.

Given $\alpha$, we will try to minimise $\beta$ (i.e. to maximise 1-$\beta$, statistical power), as much as possible.

What does statistical power depend on?

* Significance criterion ($\alpha$): fixed *a priori*
* **Effect size of interest**: defined by experimenter based on theoretical or empirical literature.
* Type of analysis: defined by experimenter.

Lets guess our power!

## Controlling our error rates | Type II error rate

Assume there **is** an effect. The probability of (incorrectly) not rejecting $H_0$ is $\beta$.

Given $\alpha$, we will try to minimise $\beta$ (i.e. to maximise 1-$\beta$, statistical power), as much as possible.

What does statistical power depend on?

* Significance criterion ($\alpha$): fixed *a priori*
* **Effect size of interest** (ES): defined by experimenter based on theoretical or empirical literature.
* Type of analysis: defined by experimenter.
* **Sample size** (N)

Lets guess our power!

## Our data

**Visual word paradigm** task

Two objects are presented, one of them is named.

How long do **monolinguals** vs. **bilinguals** look to the correct (target) object?

Independent *t*-test comparing monolinguals' and bilinguals' looking time to target.


## Power analysis | *A priori*

Given $\alpha$, *ES*, and having decided our analysis, what *N* do we need to achieve given desired power?

We know...

* Significance criterion ($\alpha$): 0.05.
* Desired stat. power (1-$\beta$): 0.80.
* Analysis: independent two-samples *t*-test

We don't know...

* Effect size (*ES*): How big is my effect size of interest? NOT THE ONE I FOUND ONCE I HAVE PERFORMED THE ANALYSIS! Our best guess is a Cohen's *d* of 0.6.

We want to calculate:

* Sample size (*N*)

## Power analysis | *A priori*

```{r power_gaze, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
pwr::pwr.t.test(
  sig.level = 0.05,          # significance criterion (alpha)
  d = 0.6,                   # effect size of interest (ES)
  type = "two.sample",       # two independent samples
  alternative = "two.sided", # both BL>ML and ML>BL can lead to rejection of H0
  power = 0.80               # desired statistical power
)
```

```{r power_gaze_plot, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
pwr::pwr.t.test(
  sig.level = 0.05,          # significance criterion (alpha)
  d = 0.6,                   # effect size of interest (ES)
  type = "two.sample",       # two independent samples
  alternative = "two.sided", # both BL>ML and ML>BL can lead to rejection of H0
  power = 0.80               # desired statistical power
) %>%
  pwr::plot.power.htest(.) # show power curve
```

## Power analysis | *A posteriori*

Our data:

<br>

```{r data_gaze, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
data <- data.frame(
  participant = paste0("id", 1:80),
  group = factor(rep(c("0", "1"), each = 40), labels = c("Monolinguals", "Bilinguals")),
  looking_time = c(rnorm(n = 40, mean = 1000, sd = 300),
                   rnorm(n = 40, mean = 1200, sd = 300))
) %>%
  mutate(
    group = fct_inorder(group)
  )

kable(rbind(head(data), tail(data)), digits = 2, col.names = c("Participant", "Group", "Looking time (ms)"), row.names = FALSE, align = "c")

```

<br>

```{r summary_gaze, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
summary <- data %>%
  group_by(group) %>%
  summarise(n = n(),
            mean = mean(looking_time),
            sd = mean(looking_time),
            sem = sd/sqrt(n))

kable(summary, digits = 2, col.names = c("Group", "n", "Mean (ms)", "SD", "SEM"), row.names = FALSE, align = "c")
```

<br>

```{r plot_gaze, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
ggplot(data, aes(x = group, y = looking_time, colour = group)) +
  geom_violin(fill = "transparent", size = 0.75, show.legend = FALSE) +
  geom_jitter(size = 4, alpha = 0.7, width = 0.1, show.legend = FALSE) +
  geom_boxplot(aes(middle = mean(looking_time)), varwidth = 0.1, fill = "transparent", colour = "black", width = 0.2, size = 0.75) +
  labs(x = "Group", y = "Looking time (ms)") + 
  scale_color_few(palette = "Medium") +
  theme(
    text = element_text(size = 12),
    axis.text = element_text(colour = "black")
  )
```

<br>

**Significance testing**:

```{r t.test, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
t.test(
  formula = looking_time ~ group, # specify response variable and predictors
  data = data                     # specify table with data
)

```

Effect size: Cohen's *d* = $t\times\sqrt{\frac{1}{n_1}+\frac{1}{n_2}}$

```{r effect_size, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
t.test <- t.test(
  formula = looking_time ~ group, # specify response variable and predictors
  data = data                     # specify table with data
)
es <- t.test$statistic/sqrt(40)

es
```

Our *t*-test yields a *p*-value lower than our $\alpha$ (0.5).

We reject the null hypothesis of equal looking times in monolinguals and bilinguals.

**What is the power of this test?**

* We can't get more participants
* We still want to keep a decent power!

*What is the lowest effect size our test would be able to detect?*

Get critical *t*-value: `t_critical = abs(qt(p = alpha, df = n-2)/sqrt(2))`

```{r t_critical, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

abs(qt(p = 0.05/2, df = (40-2)))

```

Get critical Cohen's *d*: `d_critical = t_critical/sqrt(n)`

```{r d_critical, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

abs(qt(p = 0.05/2, df = (40-2)))/sqrt(40)

```

# Simulation-based power analysis

## Same experiment over 1000 repetitions

```{r rep_gaze, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
data_gaze_big <-
  vector("list", length = 1000) %>%
  set_names(paste0("experiment", 1:1000)) %>%
  map(
    ~data.frame(
      participant = paste0("id", 1:80),
      group = factor(rep(c("0", "1"), each = 40), labels = c("Monolinguals", "Bilinguals")),
      looking_time = c(rnorm(n = 40, mean = 1000, sd = 300),
                       rnorm(n = 40, mean = 1200, sd = 300))
    )
  )

summary_gaze_big <- data_gaze_big %>%
  map(~group_by(., group)) %>%
  map(~summarise(.,
                 n = n(),
                 mean = mean(looking_time),
                 sd = sd(looking_time),
                 sem = sd/sqrt(n))) %>%
  bind_rows(., .id = "experiment")
```

```{r rep_gaze_plot_big, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
ggplot(summary_gaze_big, aes(x = group, y = mean, colour = group)) +
  geom_jitter(width = 0.1, alpha = 0.7, size = 1) +
  stat_summary(fun.y = mean, geom = "point", size = 4, colour = "black") +
  labs(x = "Group", y = "Mean looking time (ms)",
       "Mean looking time over 1000 experiments") +
  scale_colour_few(palette = "Medium") +
  theme(
    text = element_text(size = 12),
    axis.text = element_text(colour = "black"),
    panel.grid.major.x = element_blank(),
    legend.position = "none"
  )
```

```{r reps_ttest, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
t.test.list.big <- data_gaze_big %>%
  map(~t.test(looking_time ~ group, data = .))
t.test.big <-
  tibble(
    t = unlist(map(t.test.list.big, "statistic")),
    df = unlist(map(t.test.list.big, "parameter")),
    p = unlist(map(t.test.list.big, "p.value"))
  )
```

```{r reps_ttest_plot, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
ggplot(t.test.big, aes(x = p)) +
  geom_histogram(bins = 20, fill = "white", colour = "black") +
  labs(x = "P-value", y = "Numer of p-values across 1000 experiments",
       title = "P values across 100 simulations of the same experiment") +
  scale_x_continuous(breaks = seq(0, 1, 0.05)) +
  theme(
    text = element_text(size = 12),
    axis.text = element_text(colour = "black")
  )
```

## When the effect is small

```{r rep_gaze_medium, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
data_gaze_medium <-
  vector("list", length = 1000) %>%
  set_names(paste0("experiment", 1:1000)) %>%
  map(
    ~data.frame(
      participant = paste0("id", 1:80),
      group = factor(rep(c("0", "1"), each = 40), labels = c("Monolinguals", "Bilinguals")),
      looking_time = c(rnorm(n = 40, mean = 1000, sd = 300),
                       rnorm(n = 40, mean = 1050, sd = 300))
    )
  )

summary_gaze_medium <- data_gaze_medium %>%
  map(~group_by(., group)) %>%
  map(~summarise(.,
                 n = n(),
                 mean = mean(looking_time),
                 sd = sd(looking_time),
                 sem = sd/sqrt(n))) %>%
  bind_rows(., .id = "experiment")
```

```{r rep_gaze_plot_medium, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
ggplot(summary_gaze_medium, aes(x = group, y = mean, colour = group)) +
  geom_jitter(width = 0.1, alpha = 0.7, size = 1) +
  stat_summary(fun.y = mean, geom = "point", size = 4, colour = "black") +
  labs(x = "Group", y = "Mean looking time (ms)",
       "Mean looking time over 1000 experiments") +
  scale_colour_few(palette = "Medium") +
  theme(
    text = element_text(size = 12),
    axis.text = element_text(colour = "black"),
    panel.grid.major.x = element_blank(),
    legend.position = "none"
  )
```

```{r reps_ttest_medium, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
t.test.list.medium <- data_gaze_medium %>%
  map(~t.test(looking_time ~ group, data = .))
t.test.medium <-
  tibble(
    t = unlist(map(t.test.list.medium, "statistic")),
    df = unlist(map(t.test.list.medium, "parameter")),
    p = unlist(map(t.test.list.medium, "p.value"))
  )
```

```{r reps_ttest_plot_medium, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
ggplot(t.test.medium, aes(x = p)) +
  geom_histogram(bins = 20, fill = "white", colour = "black") +
  labs(x = "P-value", y = "Numer of p-values across 1000 experiments",
       title = "P values across 100 simulations of the same experiment") +
  scale_x_continuous(breaks = seq(0, 1, 0.05)) +
  theme(
    text = element_text(size = 12),
    axis.text = element_text(colour = "black")
  )
```

## When there is no effect

```{r rep_gaze_no, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
data_gaze_no <-
  vector("list", length = 1000) %>%
  set_names(paste0("experiment", 1:1000)) %>%
  map(
    ~data.frame(
      participant = paste0("id", 1:80),
      group = factor(rep(c("0", "1"), each = 40), labels = c("Monolinguals", "Bilinguals")),
      looking_time = c(rnorm(n = 40, mean = 1000, sd = 300),
                       rnorm(n = 40, mean = 1000, sd = 300))
    )
  )

summary_gaze_no <- data_gaze_no %>%
  map(~group_by(., group)) %>%
  map(~summarise(.,
                 n = n(),
                 mean = mean(looking_time),
                 sd = sd(looking_time),
                 sem = sd/sqrt(n))) %>%
  bind_rows(., .id = "experiment")
```

```{r rep_gaze_plot_no, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
ggplot(summary_gaze_no, aes(x = group, y = mean, colour = group)) +
  geom_jitter(width = 0.1, alpha = 0.7, size = 1) +
  stat_summary(fun.y = mean, geom = "point", size = 4, colour = "black") +
  labs(x = "Group", y = "Mean looking time (ms)",
       "Mean looking time over 1000 experiments") +
  scale_colour_few(palette = "Medium") +
  theme(
    text = element_text(size = 12),
    axis.text = element_text(colour = "black"),
    panel.grid.major.x = element_blank(),
    legend.position = "none"
  )
```

```{r reps_ttest_no, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
t.test.list.no <- data_gaze_no %>%
  map(~t.test(looking_time ~ group, data = .))
t.test.no <-
  tibble(
    t = unlist(map(t.test.list.no, "statistic")),
    df = unlist(map(t.test.list.no, "parameter")),
    p = unlist(map(t.test.list.no, "p.value"))
  )
```

```{r reps_ttest_plot_no, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
ggplot(t.test.no, aes(x = p)) +
  geom_histogram(bins = 20, fill = "white", colour = "black") +
  labs(x = "P-value", y = "Numer of p-values across 1000 experiments",
       title = "P values across 100 simulations of the same experiment") +
  scale_x_continuous(breaks = seq(0, 1, 0.05)) +
  theme(
    text = element_text(size = 12),
    axis.text = element_text(colour = "black")
  )
```